The Role of Variability in Learning Generalization: A Computational Modeling Approach

### 2024-05-28 Presentation/slides.qmd
update slides

---
title: "The Role of Variability in Learning Generalization: A Computational Modeling Approach"
page-layout: full
aliases:
  - /slides.html
css: /Assets/Style/slides.css
format: 
  html: 
    subtitle: "Dissertation Defense"
  clean-revealjs:
    #theme: moon #white
    #output-ext: revealjs.html
    # template-partials:
    #   - assets/title-slide.html
    output-file: slides_revealjs
    date: 05/28/204
    slideNumber: true
    hash-type: number # Use slide numbers in the URL hash
    center: true
    scrollable: true
    chalkboard:
      theme: whiteboard
  pptx:
    incremental: false
  hugo-md:
    include: false
    html-math-method: mathjax
    output-file: slides_hugo.md
  gfm:
    output: false
    output-file: slides_gfm.md
author:
  - name: Thomas Gorman
    url: https://tegorman13.github.io
    orcid: 0000-0001-5366-5442
    affiliations:
      - name: Indiana University
        department: Psychological & Brain Sciences, Cognitive Science
# params:
#   start_slide: 
#engine: knitr
---

@@

# Introduction

-   Investigate effects of training variability on learning and generalization
-   Focus areas: visuomotor skill learning and function learning tasks

@@

## Background and Motivation

-   Importance of training variability
-   Research gaps in visuomotor skill learning
-   Previous research in category learning, language learning, and function learning

@@

@@

:::::::::  columns
::: {.column width="60%"}
### Training
- Both training conditions complete the same number of training trials.
- Varied group has worse training performance.

@@

::: {.column width="40%"} 
::: {.fragment}
### Testing

@@

- Cases where varied training makes no difference
- Cases where more training variation results in worse outcomes
- Cases where the influence of variation interacts with some other factor
    - difficulty
    - prior knowledge
    - Frequency effects, or amount of training/learning before testing

@@

@@

-   Project 1: Visuomotor projectile launching task
-   Project 2: Visuomotor extrapolation task

# Project 1

-   Influence of varied practice in a projectile launching task
-   Experiments 1 & 2:
    -   Constant vs. varied training groups
    -   Performance metrics and transfer testing

@@

:::: {.column width="40%"}
- **Training Stage** - 200 training trials. Constant groups throws from single position. Varied group 60 trials from 2 locations. 
- **Transfer Stage** - All subjects tested from both positions they were trained, and the positions trained by other group
- **Data recorded** - For every throw, recorded the X velocity and Y velocity of ball at release
::::

@@

## Project 1 - Manipulations

@@

### Experiment 1
- Train a varied group from 2 positions, and a constant group from 1. Then test both groups, from the training positions and a novel position.

@@

::: {.column width="50%"}
### Experiment 2
- 6 constant conditions, each trained from a unique location, and a varied condition trained from 2 locations
:::

@@

## Project 1 - Results

:::: {.columns}
::: {.column width="40%"}
### Experiment 1
![](assets/fig-IGAS_Testing1-1-2.png)
:::
::: {.column width="60%"}
### Experiment 2
![](assets/fig-e2testa-1-2.png)
:::
::::

@@

## Model of Similarity between training and solutions

:::{#fig-solspace}
![](assets/sol_space.png){fig-align="left" width="500px"}

Solution space
:::

## Model of Similarity between training and solutions

:::: {.columns}
::: {.column width="70%"}
:::{#fig-fullspace}
![](assets/full_space.png){fig-align="left" width="700px"}

Full Space of Solution Throws and Missed Throws. Colored points represent coordinates of target hits, brown points are coordinates of misses
:::
:::
::: {.column width="30%"}
some text
:::
::::
------------------------------------------------------------------------

@@

- For each subject, compute similarity between training throws and each of the 6 testing locations
- Test whether similarity explains the difference in performance between Constant and Varied conditions
  - Assuming equivalent generalization (1 c for both groups) - similarity improves model fit, but does not explain group difference
  - Assume training condition influences generalization gradient (fit c separately) - now similarity does explain the difference between Constant and Varied.

@@

**Model-based**
- If variation does influence extrapolation, can a connectionist model with similarity-based activation provide a good account?
- Can our modelling framework simultaneously account for both training and testing data?
- Accounting for the full distribution of responses

@@

::: {.content-visible when-format="html"}
![](assets/htw_vid.mov){height="400" loop="true" autoplay="true"}
:::
-   A total of 156 participants from IU's online subject pool
-   Varied group trains from 3 "velocity bands", constant group from 1

@@

@@

## Project 2 - Methodology

-   A total of 156 participants from IU's online subject pool
-   Varied group trains from 3 "velocity bands", constant group from 1

![](assets/htw_task_fig.png)

## Project 2 - Methodology

::: {.content-visible unless-format="markdown"}
```{dot}
//| label: fig-design-e1
//| fig-cap: "Experiment 1 Design. Constant and Varied participants complete different training conditions."
//| fig-width: 9.5
//| fig-height: 3.5
//| echo: false  
digraph {
  graph [layout = dot, rankdir = LR]
  node [shape = rectangle, style = filled]
  data1 [label = " Varied Training \n800-1000\n1000-1200\n1200-1400", fillcolor = "#00A08A"]
  data2 [label = " Constant Training \n800-1000", fillcolor = "#FF0000"]
  Test3 [label = "    Final Test \n  Novel With Feedback  \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]
  data1 -> Test1
  data2 -> Test1
  subgraph cluster {
    label = "Test Phase \n(Counterbalanced Order)"
    Test1 [label = "Test  \nNovel Bands \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]
    Test2 [label = "  Test \n  Varied Training Bands  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
    Test1 -> Test2
  }
  Test2 -> Test3
}
```
:::

@@

## Project 2 - Experiment 2 Design

::: {.content-visible unless-format="markdown"}
```{dot}
//| label: fig-design-e2
//| fig-cap: "Experiment 2 Design. Constant and Varied participants complete different training conditions. The training and testing bands are the reverse of Experiment 1. "
//| fig-width: 10.0
//| fig-height: 4
//| fig-responsive: false
//| echo: false 
//| 
/////| column: screen-inset-right
digraph {
  graph [layout = dot, rankdir = LR]
  node [shape = rectangle, style = filled]
  data1 [label = " Varied Training \n100-300\n350-550\n600-800", fillcolor = "#00A08A"]
  data2 [label = " Constant Training \n600-800", fillcolor = "#FF0000"]
  Test3 [label = "    Final Test \n  Novel With Feedback  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
  data1 -> Test1
  data2 -> Test1
  subgraph cluster {
    label = "Test Phase \n(Counterbalanced Order)"
    Test1 [label = "Test  \nNovel Bands  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
    Test2 [label = "  Test \n  Varied Training Bands  \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]
    Test1 -> Test2
  }
  Test2 -> Test3
}
```
:::

@@

- Ordinal Feedback
- Both Reverse and Original Feedback Order

@@

## Theoretical Contributions

-   Insights into training variability effects on learning and generalization
-   Computational models applications

@@

## Practical Implications

-   Training programs design implications
-   Future research considerations

## Limitations and Future Directions

-   Study limitations
-   Future research suggestions

-   Recap of key findings
-   Final thoughts on training variability's role in learning

## Questions and Discussion

-   Open floor for questions from the committee and audience

## Future Research Directions {.smaller .r-fit-text}

::::::::: columns
::::: {.column width="50%"}
::: fragment
### Theoretical Extensions

-   Investigate the role of individual differences in training variability effects.
-   Explore the impact of different types of feedback on learning and generalization.
-   Examine the influence of task complexity on the benefits of varied training.
:::

::: fragment
### Methodological Innovations

-   Develop new experimental paradigms to test training variability in more complex tasks.
-   Utilize advanced computational models to simulate learning processes under varied conditions.
-   Implement longitudinal studies to assess long-term effects of training variability.
:::
:::::

::::: {.column width="50%"}
::: fragment
### Practical Applications

-   Design adaptive training programs that adjust variability based on learner performance.
-   Apply findings to educational settings to enhance learning outcomes.
-   Use insights to improve training protocols in professional skill development.
:::

::: fragment
### Collaborative Opportunities

-   Partner with educational institutions to test training variability in classroom settings.
-   Collaborate with industry professionals to apply research findings in real-world training programs.
-   Engage with interdisciplinary teams to explore the broader implications of training variability.
:::
:::::
:::::::::

>>>>

---
title: "The Role of Variability in Learning Generalization: A Computational Modeling Approach"
page-layout: full
aliases:
  - /slides.html
css: /Assets/Style/slides.css
format-links: [clean-revealjs, pptx,gfm]
format: 
  html: 
    subtitle: "Dissertation Defense"
  clean-revealjs:
    #theme: moon #white
    #output-ext: revealjs.html
    # template-partials:
    #   - assets/title-slide.html
    output-file: slides_revealjs
    date: 05/28/204
    slideNumber: true
    hash-type: number # Use slide numbers in the URL hash
    center: true
    scrollable: true
    chalkboard:
      theme: whiteboard
  pptx:
    incremental: false
  hugo-md:
    include: false
    html-math-method: mathjax
    output-file: slides_hugo.md
  gfm:
    output: false
    output-file: slides_gfm.md
author:
  - name: Thomas Gorman
    url: https://tegorman13.github.io
    orcid: 0000-0001-5366-5442
    affiliations:
      - name: Indiana University
        department: Psychological & Brain Sciences, Cognitive Science
# params:
#   start_slide: 
#engine: knitr
---

@@

# Learning Generalization

- Learning is often specific
- Longstanding scientific interest in how to improve generalization or transfer

@@

## Variability

- Variation during training linked to improved transfer in numerous domains 
- What does "variability mean in the context of learning interventions?"

## Types of Variability

- The number of unique items/problems experienced
- How spread out examples are

## My Dissertation Focus

- Number of unique examples
- Visuomotor skill learning and function learning

- Addressing methodological shortcomings of previous work
- Adapting cognitive models from other domains to account for results

@@

## Common Experimental Manipulations

- Varied vs. Constant

@@

:::::::::  columns
::: {.column width="55%"}
### Training
- Both training conditions complete the same number of training trials.
- Varied group has worse training performance.

@@

::: {.column width="45%"} 
::: {.fragment}
### Testing

@@

- Cases where varied training doesn't benefit generalization
- Cases where more training variation results in worse outcomes
- Cases where the influence of variation interacts with some other factor
    - difficulty
    - prior knowledge
    - Frequency effects, or amount of training/learning before testing

@@

::: footer
@schmidtSchemaTheoryDiscrete1975 @bjorkMakingThingsHard2011 @guadagnoliChallengePointFramework2004
:::

@@

### Project 1[^1]

- Visuomotor projectile launching task
- two experiments
- Beneficial effect of variability
- Instance-based similarity model

### Project 2

- Visuomotor extrapolation task
- Three experiments
- Effect of variability is null or negative
- Connectionist model (ALM) and hybrid associative & rule model (EXAM)

::: footer
[^1]: Now published in Cognitive Psychology @gormanInstancebasedModelAccount2022
:::
--

# Project 1  - Issues with previous research

- Assumptions about what is encoded
- Assumptions about the formation of abstractions 
- Aggregation issues and similarity confounds

---

## designs that seemingly overcome these issues

- Kerr & Booth 1978

::: footer
@kerrSpecificVariedPractice1978
:::

## Experiment 1

- Conceptual replication of Kerr & Booth design 
- Testing classic pattern as well

@@

:::: {.column width="40%"}
- **Training Stage** - 200 training trials with feedback. Constant groups trains from single position. Varied group practices from two positions. 
- **Transfer Stage** - All subjects tested from both positions they were trained, and the positions trained by other group
- **Data recorded** - For every throw, recorded the X velocity and Y velocity of ball at release
::::

@@

## Project 1 - Manipulations  {.smaller}

@@

### Experiment 1
- Train varied group from 2 positions, and a constant group from 1.
- Test both groups from the trained positions - and 1 novel position

![](assets/igas_e1_manip.png)

@@

::: {.column width="50%"}
### Experiment 2
- 6 distinct constant conditions

![](assets/igas_e2_manip.png)
:::

@@

## Project 1 - Experiment 1 Results

:::: {.columns}
::: {.column width="40%"}
### Experiment 1
![](assets/igas_e1_manip.png)
:::
::: {.column width="60%"}
::: fragment
### Training
![](assets/htt_e1_train.png)
:::
::: fragment
### Testing
![](assets/fig-IGAS_Testing1-1-2.png)
:::
:::
::::

## Project 1 - Experiment 2 Results

:::: {.columns}
::: {.column width="40%"}
### Experiment 1
![](assets/igas_e2_manip.png)
:::
::: {.column width="60%"}
::: fragment
### Training
![](assets/htt_e2_train.png)
:::
::: fragment
### Testing
![](assets/htt_e2_test.png)
:::
:::
::::

@@

![](assets/sol_space2.png){width="65%"}
![](assets/full_space2.png){width="50%"}
:::

::: {.column width="50%"}

### Model Steps

- each of the 6 positions has an empirical solution space
- Compute the summed similarity between training throws, and solution spaces
  - separately for each participant, and each of the testing positions
- We now have a measure of how similar training behavior was to testing solutions

### Model Definition

- $d_{i,j} = \sqrt{(x_{Train_i}-x_{Solution_j})^2 + (y_{Train_i}-y_{Solution_j})^2 }$
- $Similarity_{I,J} = \sum_{i=I}\sum_{j=J} (e^{-c^\cdot d^{p}_{i,j}})$

:::
::::

@@

:::: {.columns}
::: {.column width="70%"}

![](assets/sol_space.png)

![](assets/full_space.png)
:::
::: {.column width="30%"}
### Model Steps

- each of the 6 positions has an empirical solution space
- Compute the summed similarity between training throws, and solution spaces
  - separately for each participant, and each of the testing positions
- We now have a measure of how similar training behavior was to testing solutions

### Model Definition

- $d_{i,j} = \sqrt{(x_{Train_i}-x_{Solution_j})^2 + (y_{Train_i}-y_{Solution_j})^2 }$
- $Similarity_{I,J} = \sum_{i=I}\sum_{j=J} (e^{-c^\cdot d^{p}_{i,j}})$
:::
::::

:::: {.columns}
::: {.column width="60%"}
### Solution Space
![](assets/sol_space2.png)

### Full Space
:::{#fig-fullspace23}
![](assets/full_space2.png)

Full Space of Solution Throws and Missed Throws. Colored points represent coordinates of target hits, brown points are coordinates of misses
:::

### Model Steps

- each of the 6 positions has an empirical solution space
- Compute the summed similarity between training throws, and solution spaces
  - separately for each participant, and each of the testing positions
- We now have a measure of how similar training behavior was to testing solutions

### Model Definition

- $d_{i,j} = \sqrt{(x_{Train_i}-x_{Solution_j})^2 + (y_{Train_i}-y_{Solution_j})^2 }$
- $Similarity_{I,J} = \sum_{i=I}\sum_{j=J} (e^{-c^\cdot d^{p}_{i,j}})$
:::
::::

## Computational Model - Does Similarity Account for Effect of Variability?

- Test whether similarity explains the difference in performance between Constant and Varied conditions
  - Assuming equivalent generalization (1 c for both groups) - similarity improves model fit, but does not explain group difference
  - Assume training condition influences generalization gradient (fit c separately) - now similarity does explain the difference between Constant and Varied.

@@

**Model-based**
- If variation does influence extrapolation, can an associative learning model with similarity-based activation provide a good account?
- Can our modelling framework simultaneously account for both training and testing data?
- Accounting for the full distribution of responses

@@

::: {.content-visible when-format="html"}
![](assets/htw_vid.mov){height="400" loop="true" autoplay="true"}
:::

@@

- Target velocity presented at top of screen
- Participants attempt to "hit the wall" at the correct velocity
- Feedback during training - how many units above or below the target-band
- Only the "x velocity" component of the throw is relevant to the task

@@

## Project 2 - Experiment 1 Design {.nostretch .smaller}

:::: {.columns}
::: {.column width="60%"}
![](assets/htw_e1_manip.png){fig-align="left" fig-align="top" width="65%"}

![](assets/e1_design.png){fig-align="left" fig-align="top" width="70%"}
:::
::: {.column width="40%"}

- 156 participants included in final analysis
- Varied group trains from 3 "velocity bands", constant group from 1
:::
::::

@@

## Project 2 - Experiment 2 Design {.smaller}

:::: {.columns}
::: {.column width="60%"}
![](assets/htw_e2_manip.png){fig-align="left" fig-align="top"}

![](assets/e2_design.png){fig-align="left" fig-align="top"}
:::
::: {.column width="40%"}

<br><br><br><br>

- Training and Testing bands are in reversed order, relative to Experiment 1
- 110 participants included in final analysis

:::
::::

@@

- Ordinal Feedback
  - directional feedback indicating too high, too low, or correct
- Both Original (Experiment 1), and reverse (Experiment 2) orders included
- 195 participants included in final analysis

@@

## References
::: {#refs}
:::

::: {.content-visible when-format="html"}
## Manuscript {.scrollable .unnumbered .unlisted}

```{=html}
<iframe width="800" height="750" src="https://tegorman13.github.io/Dissertation/Manuscript/output/manuscript.html" title="Webpage example"></iframe>
```
:::

@@

### 2024-05-28 Sections/full.qmd
update slides

::: {.content-visible unless-format="markdown"}
```{dot}
//| label: fig-design-e2
//| fig-cap: "Experiment 2 Design. Constant and Varied participants complete different training conditions. The training and testing bands are the reverse of Experiment 1. "
//| fig-width: 7.0
//| fig-height: 2.2
//| fig-responsive: false
//| echo: false 
//| 
/////| column: screen-inset-right
digraph {
  graph [layout = dot, rankdir = LR]
  node [shape = rectangle, style = filled]
  data1 [label = " Varied Training \n100-300\n350-550\n600-800", fillcolor = "#FF0000"]
  data2 [label = " Constant Training \n600-800", fillcolor = "#00A08A"]
  Test3 [label = "    Final Test \n  Novel With Feedback  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
  data1 -> Test1
  data2 -> Test1
  subgraph cluster {
    label = "Test Phase \n(Counterbalanced Order)"
    Test1 [label = "Test  \nNovel Bands  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
    Test2 [label = "  Test \n  Varied Training Bands  \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]
    Test1 -> Test2
  }
  Test2 -> Test3
}
```

>>>>

::: {.content-visible unless-format="markdown"}
```{dot}
//| label: fig-design-e2
//| fig-cap: "Experiment 2 Design. Constant and Varied participants complete different training conditions. The training and testing bands are the reverse of Experiment 1. "
//| fig-width: 7.0
//| fig-height: 2.2
//| fig-responsive: false
//| echo: false 
//| 
/////| column: screen-inset-right
digraph {
  graph [layout = dot, rankdir = LR]
  node [shape = rectangle, style = filled]
  data1 [label = " Varied Training \n100-300\n350-550\n600-800", fillcolor = "#00A08A"]
  data2 [label = " Constant Training \n600-800", fillcolor = "#FF0000"]
  Test3 [label = "    Final Test \n  Novel With Feedback  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
  data1 -> Test1
  data2 -> Test1
  subgraph cluster {
    label = "Test Phase \n(Counterbalanced Order)"
    Test1 [label = "Test  \nNovel Bands  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
    Test2 [label = "  Test \n  Varied Training Bands  \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]
    Test1 -> Test2
  }
  Test2 -> Test3
}
```
:::
